Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-22 latex/main.tex
Add some tskit refs
Closes #761

Furthermore, some of the Pango X lineages above originate from
shared recombination events in the ARG (Table~\ref{tab:pango_x_lineages}).
An example of a set of ``nested'' closely related lineages consists of XZ, XAC, XAD, XAE, and XAP,
which descend from a single recombination event rather than five independent
events in the ARG (Figure~\ref{fig:subgraphs}B).
These Pango X lineages were designated around the same time.
These results highlight that the ARG can help clarify the relationships between recombinants,
allowing more precise interpretation of evidence for recombination.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{pango_x_subgraphs}
\caption{
Illustrative subgraphs in the ARG.
(A) Subgraph around the recombination node associated with the XA.
Two "Group A" recombinant samples reported by Jackson et al. \cite{Jackson2021}) are shown.
(B) Subgraph around the recombination node associated with a group of ``nested'' Pango X lineages; and
(C) Subgraph around the MRCA of the closely related non-recombinants XN and XAU.
The nodes assigned to a Pango X lineage are colored by that lineage.
The numbers inside the colored boxes indicate
the number of samples assigned to the Pango X lineage indicated by the same color.
The numbers below the vertical dashed lines indicate
the total number of samples within a collapsed clade.
For each recombination node, a genome schematic shows
the location of its inferred breakpoint interval relative to the Spike gene.
The ancestry above the parents of the recombination nodes or the MRCA is not shown for clarity.
For more detailed visualizations of these subgraphs, see Document S2.
}
\label{fig:subgraphs}
\end{figure}

% This table is generated by the notebook tab_pango_x_events
\begin{table}
\caption{
ARG events associated with Pango X lineages.
Each row corresponds to the oldest node assigned the specified
Pango lineage and the ancestor of all nodes assigned that label
(except XBB, XBB.1 and XM; see STAR methods). [MORE DETAILS]
}
\label{tab:pango_x_lineages}
\begin{footnotesize}
\begin{tabular}{lrrrllrl}
\toprule
\multicolumn{5}{l}{Type I RE event: monophyletic for pango}\\
pango & avt & left & right & left p & right p & samp & nested \\
\midrule
XC & 37 & 26768 & 27390 & \textbf{AY.29} & \textbf{B.1.1.7} & 5 & {} \\
XBR & 22 & 22034 & 22190 & \textbf{BN.3.1} & BQ.1.25 & 1 & {} \\
XA & 16 & 20411 & 21765 & \textbf{B.1.177.18} & \textbf{B.1.1.7} & 39 & {} \\
XS & 11 & 9054 & 10449 & AY.103 & \textbf{BA.1.1} & 17 & {} \\
XL & 8 & 5925 & 8393 & \textbf{BA.1.17.2} & \textbf{BA.2} & 64 & {} \\
XBG & 8 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BA.5.2} & 25 & {} \\
XQ & 7 & 4322 & 5386 & \textbf{BA.1.1.15} & \textbf{BA.2.9} & 55 & {BA.2:37,
XR:17, XAA:17, XU:1, XAM:21, XAG:6} \\
XBD & 7 & 23020 & 24620 & \textbf{BA.2.75.2} & \textbf{BA.5.2.1} & 30 & {} \\
XM & 6 & 17411 & 21595 & \textbf{BA.1.1} & \textbf{BA.2} & 26 & {BA.2:16,
XAL:3} \\
XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {} \\
XBB & 6 & 22332 & 22577 & \textbf{BA.2.10} & BM.1.1.1 & 6452 & {BA.2:1} \\
XY & 5 & 11538 & 12880 & \textbf{BA.1.1} & \textbf{BA.2} & 23 & {} \\
XF & 5 & 5387 & 6402 & \textbf{AY.4} & \textbf{BA.1} & 16 & {} \\
XBF & 5 & 5184 & 9866 & \textbf{BA.5.2.1} & CJ.1 & 185 & {} \\
XW & 4 & 2833 & 4321 & \textbf{BA.1.1.15} & \textbf{BA.2} & 32 & {} \\
XG & 4 & 5925 & 6513 & \textbf{BA.1.17} & \textbf{BA.2} & 3 & {} \\
\midrule
\multicolumn{5}{l}{Type II RE event}\\
alias & avt & left & right & left p & right p & desc & by pango \\
\midrule
Xx & 6 & 24504 & 26060 & \textbf{BA.2} & \textbf{BA.1.17.2} & 253 & {BA.2:156,
XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
XE/XH & 6 & 10448 & 11283 & \textbf{BA.1.17.2} & \textbf{BA.2} & 1156 &
{XE:1116, BA.2:37, XH:2, XAF:1} \\
XBM & 6 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BF.3} & 12 & {XBM:10,
BF.3:2} \\
XAF & 5 & 10199 & 10447 & \textbf{BA.1.1} & \textbf{BA.2} & 36 & {BA.2:35,
XAF:1} \\
XJ & 5 & 13196 & 17410 & \textbf{BA.1.17.2} & \textbf{BA.2} & 85 & {XJ:68,
BA.2:17} \\
\bottomrule
\multicolumn{5}{l}{Events closely derived from RE}\\
pango & alias & path len & time & mutations & samples \\
\midrule
XR & XQ & 3 & 11 & 4 & 17 \\
XU & XQ & 6 & 103 & 13 & 1 \\
XAM & XQ & 6 & 40 & 3 & 21 \\
XAA & XQ & 6 & 33 & 3 & 17 \\
XAG & XQ & 6 & 53 & 4 & 6 \\
XAP & Xx & 4 & 66 & 1 & 20 \\
XZ & Xx & 4 & 48 & 1 & 48 \\
XAE & Xx & 2 & 47 & 7 & 9 \\
XAF & XAF & 5 & 140 & 8 & 1 \\
XAL & XM & 3 & 65 & 2 & 3 \\
\midrule
\multicolumn{5}{l}{Non recombinant}\\
pango & muts & parent & samples & nested \\
\midrule
XP & 1 & \textbf{BA.1.1} & 45 & {} \\
XN & 1 & \textbf{BA.2} & 120 & {BA.2:1} \\
XBQ & 1 & CJ.1 & 14 & {} \\
XBK & 1 & CJ.1 & 7 & {} \\
XBE & 3 & BA.5.2 & 65 & {} \\
XB & 2 & B.1 & 192 & {} \\
XAZ & 3 & \textbf{BA.5} & 133 & {} \\
XAV & 3 & BA.5.1.24 & 13 & {} \\
XAU & 1 & BA.2 & 8 & {} \\
XAS & 1 & BA.4 & 77 & {} \\
XAN & 5 & \textbf{BA.5.1} & 7 & {} \\
XAJ & 11 & \textbf{BA.2.12} & 18 & {} \\
\bottomrule
\end{tabular}
\end{footnotesize}
\end{table}

@@

\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{recombinant_evidence.pdf}
\caption{
Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes,
colored according to their classification by rebar
(recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations
when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table \ref{tab:pango_x_lineages} are labeled starting with `X' and
those identified by Jackson et al. \cite{Jackson2021} are prefixed with a `J', e.g.
Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1).
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted
by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each robust recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns leading to XA, XBB, XBB.1, and the recombinant origin of XZ+XAC+XAD+XAE+XAP;
each recombination node is a mix of variants matching the left parent (green), the right parent (blue), and de-novo mutations (gold).}
\label{fig:averted_mutations}
\end{figure}

@@

\subsubsection*{Ancestral Recombination Graphs}
\label{sec:args}
The topology of an Ancestral Recombination Graph (ARG) can be defined 
% genome seems better than "haplotype" here, and will be clear what it means
% in the viral context?
as a set of \emph{nodes}, representing a genome that existed at some specific time,
and \emph{edges} that describe the genetic inheritance relationships between those 
genomes~\cite{Wong2024}. 
Each edge defines the parent and child nodes and a genomic interval 
over which inheritance occurs. The ``succinct tree sequence'' ARG encoding defines 
these relationships in a simple and concise tabular format, 
sufficient to represent arbitrarily complex patterns of
recombination~\cite{Wong2024} and providing the basis for a range of highly
efficient data processing
algorithms~\cite{Kelleher2016,Ralph2020,Kelleher2020coalescent,Grundler2025geographic,Lehmann2025on}.
The mature and feature-rich tskit library has interfaces in Python, C and Rust,
extensive documentation,
and is the basis of a growing software ecosystem of 
simulation~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,Tagami2024tstrait},
visualisation~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit},
inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022}
% More here? I'm sure there's others
and data analysis~\cite{Fan2022,Nowbandegani2023extremely}
tools.

@@

\subsection*{XS}
https://github.com/jeromekelleher/sc2ts-paper/issues/287
In the ARG, the XS samples appear to have originated from
two sequential recombination events.
However, the second event is likely an artifact,
as the XS sample associated with this event failed
the minimum net number of supporting loci.
In addition, this sample was sequenced using the AmpliSeq V1 kit,
which was also used to sequence eight other samples descending from the same event.
We excluded this likely artifactual event from further analysis and
consider XS to have emerged from a single recombinant origin.

\subsection*{XM}
https://github.com/jeromekelleher/sc2ts-paper/issues/284
Possibly related to deletions.

>>>>

Furthermore, some of the Pango X lineages above originate from
shared recombination events in the ARG (Table~\ref{tab:pango_x_lineages}).
An example of a set of ``nested'' closely related lineages consists of XZ, XAC, XAD, XAE, and XAP,
which descend from a single recombination event rather than five independent
events in the ARG (Figure~\ref{fig:subgraphs}B), and which we tentatively group using the unused label Xx.
These Pango X lineages were designated around the same time.
These results highlight that the ARG can help clarify the relationships between recombinants,
allowing more precise interpretation of evidence for recombination.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{pango_x_subgraphs}
\caption{
Illustrative subgraphs in the ARG.
(A) Subgraph around the recombination node associated with XA.
Two "Group A" recombinant samples reported by Jackson et al. \cite{Jackson2021}) are shown.
(B) Subgraph around the recombination node associated with a group of ``nested'' Pango X lineages: we use the label Xx as a shorthand for this related group of recombinants; and
(C) Subgraph around the MRCA of the closely related non-recombinants XN and XAU.
The nodes assigned to a Pango X lineage are colored by that lineage.
The numbers inside the colored boxes indicate
the number of samples assigned to the Pango X lineage indicated by the same color.
The numbers below the vertical dashed lines indicate
the total number of samples within a collapsed clade.
For each recombination node, a genome schematic shows
the location of its inferred breakpoint interval relative to the Spike gene.
The ancestry above the parents of the recombination nodes or the MRCA is not shown for clarity.
For more detailed visualizations of these subgraphs, see Document S2.
}
\label{fig:subgraphs}
\end{figure}

% This table is generated by the notebook tab_pango_x_events
\begin{table}
\caption{
ARG events associated with Pango X lineages.
Each row corresponds to the oldest node assigned the specified
Pango lineage and the ancestor of all nodes assigned that label
(except XBB, XBB.x and XM; see STAR methods). [MORE DETAILS]
}
\label{tab:pango_x_lineages}
\begin{footnotesize}
\begin{tabular}{lrrrllrl}
\toprule
\multicolumn{5}{l}{Type I RE event: monophyletic for pango}\\
pango & avt & left & right & left p & right p & samp & nested \\
\midrule
XC & 37 & 26768 & 27390 & \textbf{AY.29} & \textbf{B.1.1.7} & 5 & {} \\
XBR & 22 & 22034 & 22190 & \textbf{BN.3.1} & BQ.1.25 & 1 & {} \\
XA & 16 & 20411 & 21765 & \textbf{B.1.177.18} & \textbf{B.1.1.7} & 39 & {} \\
XS & 11 & 9054 & 10449 & AY.103 & \textbf{BA.1.1} & 17 & {} \\
XL & 8 & 5925 & 8393 & \textbf{BA.1.17.2} & \textbf{BA.2} & 64 & {} \\
XBG & 8 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BA.5.2} & 25 & {} \\
XQ & 7 & 4322 & 5386 & \textbf{BA.1.1.15} & \textbf{BA.2.9} & 55 & {BA.2:37,
XR:17, XAA:17, XU:1, XAM:21, XAG:6} \\
XBD & 7 & 23020 & 24620 & \textbf{BA.2.75.2} & \textbf{BA.5.2.1} & 30 & {} \\
XM & 6 & 17411 & 21595 & \textbf{BA.1.1} & \textbf{BA.2} & 26 & {BA.2:16,
XAL:3} \\
XBH & 6 & 15452 & 22001 & \textbf{BA.2.1} & BA.2.75.2 & 6 & {} \\
XBB & 6 & 22332 & 22577 & \textbf{BA.2.10} & BM.1.1.1 & 6452 & {BA.2:1} \\
XY & 5 & 11538 & 12880 & \textbf{BA.1.1} & \textbf{BA.2} & 23 & {} \\
XF & 5 & 5387 & 6402 & \textbf{AY.4} & \textbf{BA.1} & 16 & {} \\
XBF & 5 & 5184 & 9866 & \textbf{BA.5.2.1} & CJ.1 & 185 & {} \\
XW & 4 & 2833 & 4321 & \textbf{BA.1.1.15} & \textbf{BA.2} & 32 & {} \\
XG & 4 & 5925 & 6513 & \textbf{BA.1.17} & \textbf{BA.2} & 3 & {} \\
\midrule
\multicolumn{5}{l}{Type II RE event}\\
alias & avt & left & right & left p & right p & desc & by pango \\
\midrule
Xx & 6 & 24504 & 26060 & \textbf{BA.2} & \textbf{BA.1.17.2} & 253 & {BA.2:156,
XZ:48, XAP:20, XAC:18, XAE:9, XAD:2} \\
XE/XH & 6 & 10448 & 11283 & \textbf{BA.1.17.2} & \textbf{BA.2} & 1156 &
{XE:1116, BA.2:37, XH:2, XAF:1} \\
XBM & 6 & 22305 & 22917 & \textbf{BA.2.76} & \textbf{BF.3} & 12 & {XBM:10,
BF.3:2} \\
XAF & 5 & 10199 & 10447 & \textbf{BA.1.1} & \textbf{BA.2} & 36 & {BA.2:35,
XAF:1} \\
XJ & 5 & 13196 & 17410 & \textbf{BA.1.17.2} & \textbf{BA.2} & 85 & {XJ:68,
BA.2:17} \\
\bottomrule
\multicolumn{5}{l}{Events closely derived from RE}\\
pango & alias & path len & time & mutations & samples \\
\midrule
XR & XQ & 3 & 11 & 4 & 17 \\
XU & XQ & 6 & 103 & 13 & 1 \\
XAM & XQ & 6 & 40 & 3 & 21 \\
XAA & XQ & 6 & 33 & 3 & 17 \\
XAG & XQ & 6 & 53 & 4 & 6 \\
XAP & Xx & 4 & 66 & 1 & 20 \\
XZ & Xx & 4 & 48 & 1 & 48 \\
XAE & Xx & 2 & 47 & 7 & 9 \\
XAF & XAF & 5 & 140 & 8 & 1 \\
XAL & XM & 3 & 65 & 2 & 3 \\
\midrule
\multicolumn{5}{l}{Non recombinant}\\
pango & muts & parent & samples & nested \\
\midrule
XP & 1 & \textbf{BA.1.1} & 45 & {} \\
XN & 1 & \textbf{BA.2} & 120 & {BA.2:1} \\
XBQ & 1 & CJ.1 & 14 & {} \\
XBK & 1 & CJ.1 & 7 & {} \\
XBE & 3 & BA.5.2 & 65 & {} \\
XB & 2 & B.1 & 192 & {} \\
XAZ & 3 & \textbf{BA.5} & 133 & {} \\
XAV & 3 & BA.5.1.24 & 13 & {} \\
XAU & 1 & BA.2 & 8 & {} \\
XAS & 1 & BA.4 & 77 & {} \\
XAN & 5 & \textbf{BA.5.1} & 7 & {} \\
XAJ & 11 & \textbf{BA.2.12} & 18 & {} \\
\bottomrule
\end{tabular}
\end{footnotesize}
\end{table}

@@

\begin{figure}[h]
\centering
\includegraphics[width=1\linewidth]{recombinant_evidence.pdf}
\caption{
Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes,
colored according to their classification by rebar
(recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations
when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table \ref{tab:pango_x_lineages} are labeled starting with `X' and
those identified by Jackson et al. \cite{Jackson2021} are prefixed with a `J', e.g.
Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1).
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted
by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each robust recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns leading to XA, Xx (the origin of XZ+XAC+XAD+XAE+XAP), XBB, and an undesignated recombination event under XBB, which we here refer to as  XBB.x; See Document S1.XXX for a description of these SnpIt-like\cite{xxx} plots, in which
each recombination node is a mix of variants matching the left parent (green), the right parent (blue), and de-novo mutations (gold).}
\label{fig:averted_mutations}
\end{figure}

@@

\subsubsection*{Ancestral Recombination Graphs}
\label{sec:args}
The topology of an Ancestral Recombination Graph (ARG) can be defined 
% genome seems better than "haplotype" here, and will be clear what it means
% in the viral context?
as a set of \emph{nodes}, representing a genome that existed at some specific time,
and \emph{edges} that describe the genetic inheritance relationships between those 
genomes~\cite{Wong2024}. 
Each edge defines the parent and child nodes and a genomic interval 
over which inheritance occurs. The ``succinct tree sequence'' ARG encoding defines 
these relationships in a simple and concise tabular format, 
sufficient to represent arbitrarily complex patterns of
recombination~\cite{Wong2024} and providing the basis for a range of highly
efficient 
algorithms~\cite{Kelleher2016,Ralph2020,Kelleher2020coalescent,Grundler2025geographic,Lehmann2025on}.
The mature and feature-rich tskit library has interfaces in Python, C and Rust,
extensive documentation,
and is the basis of a growing software ecosystem of 
simulation~\cite{Baumdicker2022,Haller2019,Petr2023,Tsambos2023link,Tagami2024tstrait},
visualisation~\cite{Karthikeyan2025tsbrowse,Kitchens2025tskit},
inference~\cite{Kelleher2019,Wohns2022,Mahmoudi2022,Martin2025model,Pieszko2025detecting}
% More here? I'm sure there's others
and data analysis~\cite{Fan2022,Nowbandegani2023extremely,Fritze2024forest}
tools.

@@

\subsubsection*{XS}
https://github.com/jeromekelleher/sc2ts-paper/issues/287
In the ARG, the XS samples appear to have originated from
two sequential recombination events.
However, the second event is likely an artifact,
as the XS sample associated with this event failed
the minimum net number of supporting loci.
In addition, this sample was sequenced using the AmpliSeq V1 kit,
which was also used to sequence eight other samples descending from the same event.
We excluded this likely artifactual event from further analysis and
consider XS to have emerged from a single recombinant origin.

\subsubsection*{XM}
https://github.com/jeromekelleher/sc2ts-paper/issues/284
Possibly related to deletions.
